Generalized Bradley-Terry models

### 2025-03-21 inst/jss_hyper3_real.Rnw
Equation not equation

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3_jss} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3_jss} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR_jss}.

@@

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
lam <- seq(from=0.9, to=3, len=howmany)
logLike <- sapply(lam,f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

\begin{equation}\label{asuccb}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \lambda p_c + p_d + p_e}\cdot
\frac{\lambda p_c}{\lambda p_c + p_d + p_e}\cdot
\frac{p_d}{p_d + p_e}
\end{equation}
%
In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
this point, no competitor has finished and cheering effects are
absent.  The second term reflects the fact that competitors $b$ and
$c$ are supported by competitor $a$, who has by this point finished
the race and is supporting his teammates.

>>>>

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
Equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3_jss} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3_jss} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR_jss}.

@@

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache=TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda),give=1)$`Log-likelihood`}
lam <- seq(from=0.9, to=3, len=howmany)
logLike <- sapply(lam,f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

\begin{equation}\label{asuccb}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \lambda p_c + p_d + p_e}\cdot
\frac{\lambda p_c}{\lambda p_c + p_d + p_e}\cdot
\frac{p_d}{p_d + p_e}
\end{equation}
%
In Equation \ref{asuccb}, the first term is standard Plackett-Luce: at
this point, no competitor has finished and cheering effects are
absent.  The second term reflects the fact that competitors $b$ and
$c$ are supported by competitor $a$, who has by this point finished
the race and is supporting his teammates.
